Separation of an upwelling current bounding the Juan de Fuca Eddy

### 2023-01-24 PW13Paper.tex
Version to SA, SNW after revisions

Comprehensive water column observations of temperature, salinity, and oxygen on the productive southern Vancouver Island shelf offer a unique characterization of processes affecting exchange on the shelf.  A semi-permanent cyclonic recirculation (the Juan de Fuca Eddy) occupies the region in the lee of a bank where the shelf widens abruptly.  The observations indicate that the water in this Eddy is a mixture of offshore water and water from a buoyant coastal current.  This water is well-mixed along a mixing line in temperature-salinity space, though it retains stratification, and is either rapidly mixed (high mixing rate $\kappa$), or has a long residence time (large $\tau$), such that the mixing length $(\kappa \tau)^{1/2} \sim 50$ ---  $100\ \mathrm{m}$.  There is a sharp temperature-salinity front on the offshore side of this well-mixed water. The front is no more than 1-km wide, and has no sign of instabilities. Previous work had hypothesized the role of a spur canyon incising the shelf in feeding water to the Eddy; little evidence of this was found in the observations, however, these were collected during a relaxation and reversal of upwelling winds, which may have slowed the flow of water into the Eddy.  Tidally resolved observations in the spur canyon also show very strong hydraulic flows in the cross-canyon direction accompanied by very high localized mixing rates in 50-m hydraulic jumps, possibly exceeding $\kappa \sim 1 \ \mathrm{m^2s^{-1}}$.

Upstream of the Eddy, there is an along-shelf current flowing equatorward with partially mixed water properties compared to the well-mixed water in the Eddy.  The shelf current was observed to separate from the shelf in the lee of a bank, and was replaced by a distinct offshore water mass.  The separation event can be seen in sea-surface temperatures in satellite images as a tongue of water that is ejected offshore.  The cause of this offshore separation event is not determined, but possibly is catalyzed by separation of the current from the abrupt underwater bank at the poleward end of the study region.

@@

The southern Vancouver Island continental shelf is very biologically productive due to high nutrient input from the Strait of Juan de Fuca and Salish Sea estuarine system and substantial cross-shelf transport due to the complicated topography.  The estuarine water flows poleward, hugging the coast as the Vancouver Island Coastal Current and, in the summer, shelf water flows equatorward.  Trapped between these two currents is the Juan de Fuca Eddy, a stagnant region of water that occupies the shelf where it abruptly widens just north of the Strait of Juan de Fuca.

Here we present intensive sampling of the Juan de Fuca Eddy region. The observations show that below the surface mixed layer, the water in the Eddy is low in oxygen, and has undergone substantial vertical and lateral mixing. In contrast to previous literature we find that the low oxygen in the eddy is likely because of respiration rather than being pulled from low-oxygen water in the California Undercurrent.

The observations also show a remarkable flow separation of the equatorward shelf current.  The current is seen to detach and is pushed offshore. Such events are readily seen in satellite imagery, but our observations indicate that the separation extends the depth of the water column on the shelf, and that this separation may be partially driven by the local bathymetry.  The separation is a very strong cross-shelf exchange event, and likely transports substantial nutrient-rich coastal water offshore to drive productivity in the deeper ocean adjacent to the continental slope.

@@

Cross-shelf exchange is important to the health and productivity of continental shelf regions, allowing for offshore oxygenated water to be exchanged with nutrient-rich, but oxygen-depleted, nearshore water.  Cross-shelf transport usually requires ageostrophic flow, since geostrophically balanced flow will tend to follow topographic contours, often providing a barrier to lateral exchange \cite{brink16}.  Mechanisms of cross-shelf exchange include internal waves and instabilities in shelf-break fronts.  However, these mechanisms are often smaller in magnitude to intermittent three-dimensional exchange, driven by eddying, and often catalyzed by topographic irregularity \cite{barthetal00}.

Here we present detailed \emph{in-situ} observations from the southern Vancouver Island shelf, collected in summer 2013, and sampled by a rapid profiling vehicle equipped with a CTD and oxygen sensor, supplemented by traditional hydrographic surveys bracketing the high resolution observations by a month. The shelf has complicated bathymetry (\fref{fig:LocMapBoth}), with a simple shelf geometry poleward of our study site, but widening in the lee of La Perouse Bank, and incised by the Juan de Fuca canyon, equatorward.  Water from the Strait of Juan de Fuca flows poleward as a buoyant current that hugs the coast \cite{thomsonetal89, hickeyetal91}, while shelf water flows equatorward in the summer under upwelling conditions, both forced by local winds and via tele-connections with the long, homogenous shelves equatorward off Washington and Oregon \cite{hickeyetal91,thomsonkrassovski15,engidaetal16}.  Trapped between these two currents is a region of relatively homogenous water that has been termed the Juan de Fuca Eddy \cite{freelanddenman82,freelandmcintosh89,foremanetal07,macfadyenhickey10}, denoted ``\Eddy'' below.

A goal of our study was to understand how the \Eddy\ persists and how it exchanges water properties with offshore water.  We focus on water deeper than $50 \mathrm{m}$, below the summer mixed layer, because this water is trackable with water mass properties, whereas the \Eddy\ is often studied from the point of view of the surface circulation \citep[e.g.][]{macfadyenhickey10}.  Past studies have found low oxygen concentrations in the \Eddy, and inferred that water is upwelled from the California Undercurrent from as deep as 400 m \cite{freelanddenman82,deweycrawford88}.  This inference was based on the assumption that oxygen is conservative, and hence water in the eddy had to come from  the oxygen minimum zone found further offshore \cite{mackasetal87}.  This lead to hypotheses that perhaps the \Eddy\ was fed by waters being drawn up a spur canyon (called the Spur Canyon) via ageostrophic transport from offshore to onshore due to low pressure in the \Eddy\ center \cite{weaverhsieh87}.  Below, we will argue that there is no evidence of such transport and that oxygen is likely low because of local consumption due to respiration.

A second goal of our study was to better understand cross-shore exchange between shelf and offshore water.  Cross-shore transport is supplied by Ekman layers in simple geometries under wind forcing. However, in many locations there is also evidence of filaments and wholesale separation of shelf currents into the deep ocean.  Along the Vancouver Island shelf, large filaments or instabilities of the coastal current have been inferred using satellite images \cite{ikedaemery84,thomsongower98}.  Direct observations of such filaments have been made further to the south off Cape Blanco \cite{barthetal00}, and off other coasts \cite{relvasbarton05}. Net cross-shore transport of nutrients and chlorophyll have been found in hydrographic surveys off Vancouver Island \cite{mackasyelland99}, and associated with mesoscale features in geostrophic velocities and in satellite observations.  The mechanisms driving such separations are poorly understood, but hypotheses include coastal hydraulics leading to along-shore trapping of coastal waves \cite{dalebarth01},  wind stress curl variations \cite{castelaobarth07}, induced relative vorticity due to stretching of parcels that overshoot their initial isobaths due to a sudden change in downstream bathymetry \citep{dasaro88}, and the non-linear breaking of large-scale meanders due to baroclinic instability between the wind-driven current and the California Undercurrent \cite{ikedaetal84, batteen97}.

@@

Below, we focus on finescale surveys carried out from 2013-08-21 to 2013-08-30.  Data were collected from the \emph{R/V Falkor} with an ODIM Brooke-Ocean Moving Vessel Profiler (MVP).  The MVP was equipped with an AML Oceanographic CTD, and a Rinko Oxygen sensor with a 7-s response time foil.  The MVP profiled to depths of 200 m or to within 5 m of the seafloor, whichever was shallower, and dropped at a speed of approximately $3\ \mathrm{m\,s^{-1}}$.  Data collection took place while the ship cruised at speeds between 5 and 8 kts, usually at around 6 kts, to enable fine horizontal spacing of the casts, with typical spacing  of 800 m in deep water, and less in water shallower than 200 m.  Data is reported for the downcast, which mostly follows a vertical path.  The rapid speed of the profiling makes the oxygen measurements somewhat coarse, and probably biased due to the phase lag of the sensor, so we treat these qualitatively in this paper.

@@

Winds during the cruises were typical for the west coast of Vancouver Island, with upwelling favorable winds during July and early August (\fref{fig:LaPeWind}).  During the finescale survey, and for the week previous, the winds were intermittently downwelling favorable.  Note that this locale is strongly affected by coastally trapped waves from further south, so doming of near-bottom isopycnals often persists despite local wind forcing, and takes a finite amount of time to spin down \cite{thomsonkrassovski15, engidaetal16}.

@@

Hydrographic sections along the LB and LC lines highlight summer conditions on the Southern Vancouver Island Shelf and indicate some of the features we are focusing on in this paper (\Fref{fig:LaPerouse2013Ctd}).  During both surveys, and along both lines, there is clear evidence of upwelling, with the $26.4\ \mathrm{kg\,m^{-3}}$ isopycnal reaching from 130 m depth offshore to shallower than 85 m over the shelf.  This upwelled water tends to be low in oxygen and cool. Further onshore, the Vancouver Island Coastal Current hugs the coast, where isopycnals tilt back down and water is warmer than offshore.

@@

The water upstream of the \Eddy\ region, along the LC line, is less mixed than the \Eddy\ water (\fref{fig:LaPerouse2013Ctd}, top three rows).  It does not show as much drawdown of oxygen, and is warmer, at least offshore of La Perouse bank ($X \approx 5\ \mathrm{km}$).  Based on these properties, offshore water does not appear to make it over La Perouse bank into the basin onshore, or if it does, it does so intermittently. Note that the water along the $26.4\ \mathrm{kg\,m^{-3}}$ isopycnal becomes cooler and lower in oxygen where it intersects the shelf, consistent with both enhanced mixing near the shelf and with drawdown of oxygen by respiration.

@@

We use a spice anomaly defined as relative to a straight line in $\theta$--$S$ space (\fref{fig:LaPerouse2013TS}) representing an approximate mixing line for water in the \Eddy, and passsing through the points $7.75\ \mathrm{^oC}$, $30\ \mathrm{psu}$ and $6.6\ \mathrm{^oC}$, $35\ \mathrm{psu}$. Spice anomaly for a given water sample at density $\sigma_{\theta}$ is given by $\gamma = \alpha \left(T - T_0\left(\sigma_{\theta}\right)\right) + \beta \left(S - S_0\left(\sigma_{\theta}\right)\right)$ where $T_0\left(\sigma_{\theta}\right)$, where $S_0\left(\sigma_{\theta}\right)$ are the temperature and salinity along the mixing line at the same density as the water sample. The sign convention is that a positive anomaly is warmer and saltier than data along the mixing line.

@@

There are temporal changes over the summer.  Away from the surface, water has upwelled from deeper depths by September, but the offshore water maintains the same water mass characteristics through the summer.  Hugging the coast ($X > 30\, \mathrm{km}$), the Vancouver Island Coastal Current warms during the summer, and the spice anomaly goes from negative to postive.  In the \Eddy, the water stays near the mixing line, but is cooler in the spring (\fref{fig:LaPerouse2013Ctd}, LB Line, left-hand column: slightly negative spice anomaly) and warms during the summer (\fref{fig:LaPerouse2013Ctd}, LB Line, right-hand column: slightly positive spice anomaly).

Oxygen concentration in both surveys has a similar dichotomy between the \Eddy\ and offshelf water.  Water found in the \Eddy\ has oxygen concentrations $100\ \mathrm{\mu mol\, kg^{-1}}$ lower on the shelf than offshelf (\fref{fig:LaPerouse2013Ctd}, \fref{fig:LaPerouse2013TS}b).  The very deepest water ($S\approx 33.9\ \mathrm{psu}$) on the shelf shows a further $50\ \mathrm{\mu mol\, kg^{-1}}$ decrease in concetration between May and September along the LB line.

\subsection{Finescale surveys}

\subsubsection{Overview}

The finescale surveys covered most of the \Eddy\ region, with an emphasis on the offshore edge near the shelfbreak front (\fref{fig:LocMapBoth}).  For water denser than $26\ \mathrm{kg\,m^{-3}}$, there are three distinct water masses sampled on the shelf (\fref{fig:TSdensSpice}).  The first is the offshore water, which tends to be warmer, and hence has a positive spice anomaly ($\gamma \approx 0.2\ \mathrm{kg\,m^{-3}}$ along $\sigma_{\theta} = 26.4\ \mathrm{kg\,m^{-3}}$).  The second is water in the \Eddy, which during this survey was found along the straight mixing line in $\theta-S$ space (\fref{fig:TSdensSpice}, $\gamma \approx 0\ \mathrm{kg\,m^{-3}}$ along $\sigma_{\theta} = 26.4\ \mathrm{kg\,m^{-3}}$).  Between these two water masses, there is a less populous mass (\fref{fig:TSdensSpice}, $\gamma \approx 0.1\ \mathrm{kg\,m^{-3}}$ along $\sigma_{\theta} = 26.4\ \mathrm{kg\,m^{-3}}$) that we demonstrate below is found on the shelf poleward of the \Eddy.

% TODO: shallower?

@@

In contrast, the third population of partially mixed water between the \Eddy\ and offshore water (\fref{fig:TSdensSpice}) is all found poleward of the \Eddy\ region (\fref{fig:CrossSectionsSpice}a, b) as water with a weaker spice anomaly ($\gamma \approx 0.1\ \mathrm{kg\,m^{-3}}$, pink colors).  This water is generally found near the intersection of the upwelling water and La Perouse Bank, and is not as warm as offshore water, indicating some mixing with offshore water has taken place.  Of note is that this intermediate water mass appears completely absent in the sections further equatorward (\fref{fig:CrossSectionsSpice}c, d), indicating that it is either mixed away or that it is advected elsewhere; we argue below that it is advected offshore in a large-scale tongue.

@@

Oxygen saturation sections show similar trends, with the low-spice anomaly water also having low oxygen saturation, while the strong negative spice anomaly shallower than $\sigma_{\theta} = 26.4\,\mathrm{kg\,m^{-3}}$ has very high oxygen saturation; deeper water is also more oxygenated.  Of note is that compared with the coarser sampling along the LC Line (\fref{fig:LaPerouse2013Ctd}), we see that the abrupt transition between offshore and \Eddy\ water is also present in disolved oxygen.

@@

The spatial patterns are very clear when considering a map view of properties along the $\sigma_{\theta} = 26.4\,\mathrm{kg\,m^{-3}}$ isopycnal (\fref{fig:SpiceO2264}).  There is a region onshore of the south tip of La Perouse Bank (approximately 44.5\textdegree N, 125.75\textdegree W), that consists of water that is found along the mixing line (spice anomaly $\gamma \approx 0\ \mathrm{kg\,m^{-3}}$) with low oxygen saturation.  Offshore of this region is water that is very warm and salty in comparison ($\gamma > 0.15 \mathrm{kg\,m^{-3}}$), and relatively high in oxygen (saturations of approximately 60\%).  The region between these two water masses is very abrupt, and  stretches from La Perouse Bank to a shallow bank just above the Juan de Fuca canyon (48.3\textdegree N and 125.4\textdegree W).

@@

\subsubsection{Spur Canyon}

The Spur Canyon leading from the Strait of Juan de Fuca has been implicated in allowing dense water to be upwelled into the \Eddy.  The sea surface is low in the middle of the \Eddy, so it has been hypothesized that water moves up the Spur Canyon due to ageostrophic motion \cite{weaverhsieh87,freelanddenman82}.  This is difficult to infer from the observations collected here.  Three transects up the canyon indicate that deep isopycnals slope up into the canyon (\fref{fig:CanyonPropertiesSpice}, to approximately 30 km).  Continuing across the \Eddy, isopycnals are largely flat until they intersect the Vancouver Island Coastal Current ($80\ \mathrm{km}$).  The deeper isopycnals are not found in the deeper basin northeast of La Perouse bank ($70\ \mathrm{km}$),  thus the bank is a natural northern boundary of the \Eddy.

@@

Much of the modified water in the Spur Canyon appears to come from the shelf to the west, but heavily modified by tidal mixing. A repeated tidal survey over the bank on the west side of the canyon shows a strong hydraulic response during onshore flow (17:58--21:00, \fref{fig:HydraulicsCanyon}).  Dense water passes from the offshore side into the canyon, and plunges down the side wall before rebounding downstream.  Note that the tide here is largely diurnal, so this onslope flow only occurs once a day.  Given the stratification of $N\approx 6\times10^{-3} \ \mathrm{rad\,s^{-1}}$ and an overturning scale of 50 m, we might expect dissipation rates reaching $\epsilon \sim L^2N^{3} \approx 5\times10^{-4}\ \mathrm{m^2\,s^{-3}}$, which is more than three orders of magnitude higher than dissipation observed west of this location by \citeA{deweycrawford88}.  This estimate of turbulence dissipation rate implies a diapycnal diffusivity of $\kappa = \gamma \epsilon / N^2 \approx 1\ \mathrm{m^2\,s^{-1}}$, assuming a mixing efficiency of $\gamma=0.2$. Water spills over from the offshore front into the canyon during the onshore tide.  This water is rapidly mixed with surrounding water such that its strong offshore spice values are attenuated.

@@

The turbulence found on the canyon rim makes it ambiguous if there is water moving up the canyon or not.  There is a general tendency along the canyon for higher spice water to be found offshore (\fref{fig:CanyonPropertiesSpice}) but it seems likely the source of the higher spice water is from over the bank rather than water being advected up the canyon.  This tidally driven flow over the bank is the most significant source of high-spice offshore water into the eddy region identified during our surveys.

\subsubsection{Frontal survey}
\label{sec:frontsurvey}

@@

\subsection{Separation of coastal water}

>>>>

Observations of temperature, salinity, and oxygen on the southern Vancouver Island shelf show a large-scale exchange of shelf water with offshore water, just offshore of a semi-permanent recirculation.   The semi-permanent cyclonic recirculation (the Juan de Fuca Eddy) occupies a region where the shelf widens abruptly in the lee of a bank.  The water in this Eddy is a mixture of offshore water and water from a buoyant coastal current.  This water is well-mixed along a mixing line in temperature-salinity space, though it retains stratification, and is either rapidly mixed or has a long residence time.  There is a sharp temperature-salinity front on the offshore side of this well-mixed water, no more than 1-km wide, and has no sign of instabilities. The clearest evidence of cross-front transport is found during a short tidally resolved survey over a bank, and was due to flows in the cross-bank direction driving large mixing in 50-m tall hydraulic jumps. Upstream of the recirculation there is an along-shelf current flowing equatorward.  However the whole current separates from the shelf before reaching the recirculation, in the lee of a bank, and is replaced by water from offshore.  The separation event was also seen in sea-surface temperatures from satellite images as a tongue of cool coastal water that is ejected offshore.

@@

The southern Vancouver Island continental shelf is biologically productive due to high nutrient input from the Strait of Juan de Fuca and Salish Sea estuarine system and substantial cross-shelf transport due to the complicated topography.  Here we present intensive sampling of the Juan de Fuca Eddy region. The observations show that below the surface mixed layer, the water in the Eddy is low in oxygen, and has undergone substantial vertical and lateral mixing. In contrast to previous literature we find that the low oxygen in the eddy is likely because of respiration rather than being pulled from low-oxygen water in the California Undercurrent.

The observations also show a remarkable flow separation of the equatorward shelf current.  The current is seen to detach and is pushed offshore. Such events are readily seen in satellite imagery, but our observations indicate that the separation extends the depth of the water column on the shelf, and that this separation may be partially driven by the local bathymetry.  The separation is a very strong cross-shelf exchange event, and transports substantial nutrient-rich coastal water offshore to drive productivity in the deeper ocean adjacent to the continental slope.

@@

Cross-shelf exchange is important to the health and productivity of continental shelf regions, allowing for offshore oxygenated water to be exchanged with nutrient-rich, but oxygen-depleted, nearshore water.  Cross-shelf transport usually requires ageostrophic flow, since geostrophically balanced flow will tend to follow topographic contours, often providing a barrier to lateral exchange \cite{brink16}.  Mechanisms of cross-shelf exchange include internal waves and instabilities in shelf-break fronts.  However, these mechanisms can be smaller in magnitude compared to intermittent three-dimensional exchange, driven by eddying, often catalyzed by topographic irregularity \cite{barthetal00}.

Here we present detailed \emph{in-situ} observations from the southern Vancouver Island shelf, collected in summer 2013, and sampled by a rapid profiling vehicle equipped with a CTD and oxygen sensor, supplemented by traditional hydrographic surveys bracketing the high resolution observations by a month. The shelf has complicated bathymetry (\fref{fig:LocMapBoth}), with a relatively simple 50-km wide shelf poleward of our study site, that widens to over 75 km wide equatorward of La Perouse Bank.  The South Vancouver Island Shelf is this wide shelf region, characterized by a number of banks, and finally is incised on the poleward side by the Juan de Fuca canyon.   Water from the Strait of Juan de Fuca flows poleward as a buoyant current that hugs the coast \cite{thomsonetal89, hickeyetal91}, while shelf water flows equatorward in the summer both forced by local winds and via teleconnections with the long, homogenous shelves equatorward off Washington and Oregon \cite{hickeyetal91,thomsonkrassovski15,engidaetal16}.  Trapped between these two currents is a region of relatively homogenous water that has been termed the Juan de Fuca Eddy \cite{freelanddenman82,freelandmcintosh89,foremanetal07,macfadyenhickey10}, denoted ``\Eddy'' below.

A goal of our study was to understand how the \Eddy\ persists and how it exchanges water properties with offshore water.  We focus on water deeper than $50 \mathrm{m}$, below the summer mixed layer, because this water is trackable with water mass properties, even though the \Eddy\ is often studied from the point of view of the surface circulation \citep[e.g.][]{macfadyenhickey10}.  Past studies have found low oxygen concentrations in the \Eddy, and inferred that water is upwelled from the California Undercurrent from as deep as 400 m \cite{freelanddenman82,deweycrawford88}.  This inference was based on the assumption that oxygen is conservative, and hence water in the \Eddy\ had to come from  the oxygen minimum zone found further offshore \cite{mackasetal87}.  This lead to hypotheses that perhaps the \Eddy\ was fed by waters being drawn up a spur canyon (called the Spur Canyon) via ageostrophic transport from offshore to onshore due to low pressure in the \Eddy\ center \cite{weaverhsieh87}.  Below, we will argue that there is no evidence of such transport and that oxygen is likely low because of local consumption due to respiration.

A second goal of our study was to better understand cross-shore exchange between shelf and offshore water.  Cross-shore transport is supplied by Ekman layers in simple geometries under wind forcing. However, in many locations there is also evidence of eddies and filaments driving wholesale separation of shelf currents into the deep ocean.  Along the Vancouver Island shelf, large filaments or instabilities of the coastal current have been inferred using satellite images \cite{ikedaemery84,thomsongower98}.  Direct observations of such filaments have been made further to the south off Cape Blanco \cite{barthetal00}, and off other coasts \cite{relvasbarton05}. Net cross-shore transport of nutrients and chlorophyll have been found in hydrographic surveys off Vancouver Island \cite{mackasyelland99}, and associated with mesoscale features in geostrophic velocities and in satellite observations.  The mechanisms driving such separations are poorly understood, but hypotheses include coastal hydraulics leading to along-shore trapping of coastal waves \cite{dalebarth01},  wind stress curl variations \cite{castelaobarth07}, induced relative vorticity due to stretching of parcels that overshoot their initial isobaths due to a sudden change in downstream bathymetry \citep{dasaro88}, and, at this location, non-linear breaking of large-scale meanders due to baroclinic instability between the wind-driven current and the California Undercurrent \cite{ikedaetal84, batteen97}.

@@

We focus on finescale surveys carried out between these hydrographic surveys, from 2013-08-21 to 2013-08-30.  Data were collected from the \emph{R/V Falkor} with an AML Oceanographic Moving Vessel Profiler (MVP).  Data was collected analogously to data collected during similar field campaigns \cite{klymaketal15a, klymaketal16, dasaroetal18}. The MVP was equipped with an AML Oceanographic CTD, and a Rinko Oxygen sensor with a 7-s response time foil.  The MVP profiled to depths of 200 m or to within 5 m of the seafloor, whichever was shallower, and dropped at a speed of approximately $3\ \mathrm{m\,s^{-1}}$.  Data collection took place while the ship cruised at speeds between 5 and 8 kts, usually at around 6 kts, to enable fine horizontal spacing of the casts, with typical spacing  of 800 m in deep water, and less in water shallower than 200 m.  Data is reported for the downcast, which mostly follows a vertical path.  The rapid speed of the profiling makes the oxygen measurements somewhat coarse, and probably biased due to the phase lag of the sensor, so we treat these qualitatively in this paper.

@@

Winds during the cruises were typical for the west coast of Vancouver Island, with equatorward upwelling-favorable winds during July and early August (\fref{fig:LaPeWind}).  During the finescale survey, and for the week previous, the winds were intermittently downwelling favorable.  Note that this locale is strongly affected by coastally trapped waves from further south, so doming of near-bottom isopycnals often persists despite local wind forcing, and takes a finite amount of time to spin down \cite{thomsonkrassovski15, engidaetal16}.

@@

Hydrographic sections along the LB and LC hydrographic lines highlight summer conditions on the Southern Vancouver Island Shelf and indicate some of the features we are focusing on in this paper (\Fref{fig:LaPerouse2013Ctd}).  During both surveys, and along both lines, there is clear evidence of upwelling, with the $26.4\ \mathrm{kg\,m^{-3}}$ isopycnal reaching from 130 m depth offshore to shallower than 85 m over the shelf.  This upwelled water tends to be low in oxygen and cool. Further onshore, the Vancouver Island Coastal Current hugs the coast, where isopycnals tilt down towards the shore,and water is warmer than offshore.

@@

The water upstream of the \Eddy\ region, along the LC line, is less mixed than the \Eddy\ water (\fref{fig:LaPerouse2013Ctd}, top three rows).  It does not show as much drawdown of oxygen, and is warmer, at least offshore of La Perouse bank ($X \approx 5\ \mathrm{km}$).  Based on these properties, offshore water does not appear to make it over La Perouse bank into the onshore basin, or if it does, it does so intermittently and with substantial mixing. Note that the water along the $26.4\ \mathrm{kg\,m^{-3}}$ isopycnal becomes cooler and lower in oxygen where it intersects the shelf, consistent with both enhanced mixing near the shelf and with drawdown of oxygen by respiration.

@@

We use a spice anomaly defined as relative to a straight line in $\theta$--$S$ space (\fref{fig:LaPerouse2013TS}) representing an approximate mixing line for water in the \Eddy, and passing through the points $7.75\ \mathrm{^oC}$, $30\ \mathrm{psu}$ and $6.6\ \mathrm{^oC}$, $35\ \mathrm{psu}$. Spice anomaly for a given water sample at density $\sigma_{\theta}$ is given by $\gamma = \alpha \left(T - T_0\left(\sigma_{\theta}\right)\right) + \beta \left(S - S_0\left(\sigma_{\theta}\right)\right)$ where $T_0\left(\sigma_{\theta}\right)$, where $S_0\left(\sigma_{\theta}\right)$ are the temperature and salinity along the mixing line at the same density as the water sample. The sign convention is that a positive anomaly is warmer and saltier than data along the mixing line.

@@

There are temporal changes over the summer.  Away from the surface, water has upwelled from deeper depths by September, but the offshore water maintains the same water mass characteristics through the summer.  Hugging the coast ($X > 30\, \mathrm{km}$), the Vancouver Island Coastal Current warms during the summer, and the spice anomaly goes from negative to positive.  In the \Eddy, the water stays near the mixing line, but is cooler in the spring (\fref{fig:LaPerouse2013Ctd}, LB Line, left-hand column: slightly negative spice anomaly) and warms during the summer (\fref{fig:LaPerouse2013Ctd}, LB Line, right-hand column: slightly positive spice anomaly).

Oxygen concentration in both surveys has a similar dichotomy between the \Eddy\ and off-shelf water.  Water found in the \Eddy\ has oxygen concentrations $100\ \mathrm{\mu mol\, kg^{-1}}$ lower on the shelf than off-shelf (\fref{fig:LaPerouse2013Ctd}, \fref{fig:LaPerouse2013TS}b).  The very deepest water ($S\approx 33.9\ \mathrm{psu}$) on the shelf shows a further $50\ \mathrm{\mu mol\, kg^{-1}}$ decrease in concetration between May and September along the LB line.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Finescale surveys}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Overview}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The finescale surveys covered most of the \Eddy\ region, with an emphasis on the offshore edge near the shelfbreak front (\fref{fig:LocMapBoth}).  For water denser than $26\ \mathrm{kg\,m^{-3}}$, there are three distinct water masses sampled on the shelf (\fref{fig:TSdensSpice}).  The first is offshore water, which tends to be warmer, and hence has a positive spice anomaly ($\gamma \approx 0.2\ \mathrm{kg\,m^{-3}}$ along $\sigma_{\theta} = 26.4\ \mathrm{kg\,m^{-3}}$).  The second is water in the \Eddy, which during this survey was found along the straight mixing line in $\theta-S$ space (\fref{fig:TSdensSpice}, $\gamma \approx 0\ \mathrm{kg\,m^{-3}}$ along $\sigma_{\theta} = 26.4\ \mathrm{kg\,m^{-3}}$).  Between these two water masses, there is a less populous mass (\fref{fig:TSdensSpice}, $\gamma \approx 0.1\ \mathrm{kg\,m^{-3}}$ along $\sigma_{\theta} = 26.4\ \mathrm{kg\,m^{-3}}$) that we demonstrate below is found on the shelf poleward of the \Eddy.

@@

In contrast, the third population of partially mixed water between the \Eddy\ and offshore water (\fref{fig:TSdensSpice}) is all found poleward of the \Eddy\ region (\fref{fig:CrossSectionsSpice}a, b) as water with a weaker spice anomaly ($\gamma \approx 0.1\ \mathrm{kg\,m^{-3}}$, pink colors).  This water is  not as warm as offshore water, indicating some mixing with offshore water has taken place.  Of note is that this intermediate water mass appears completely absent in the sections further equatorward (\fref{fig:CrossSectionsSpice}c, d), indicating that it is either mixed away or that it is advected elsewhere; we argue below that it is advected offshore in a large-scale tongue.

@@

Oxygen saturation sections show that the deep water in the  \Eddy\ has very low oxygen saturation (\fref{fig:CrossSectionsO2}c, d) compared to surrounding water (though again, we caution against the quality of these saturation values from such a fast profiling instrument with a slow response time).  More oxygenated water is found offshore, and the oxygen deficit is not as strong in the poleward sections (\fref{fig:CrossSectionsO2}a, b).  The T/S compensated front also shows an abrupt transition from the \Eddy\ water and the offshore. Note the excellent correlation between oxygen saturation and spice anomaly in \fref{fig:CrossSectionsSpice}.

@@

The spatial patterns are very clear when considering a map view of properties along the $\sigma_{\theta} = 26.4\,\mathrm{kg\,m^{-3}}$ isopycnal (\fref{fig:SpiceO2264}).  There is a region onshore of the south tip of La Perouse Bank (approximately 44.5\textdegree N, 125.75\textdegree W), that consists of water that is found along the mixing line (spice anomaly $\gamma \approx 0\ \mathrm{kg\,m^{-3}}$) that also has low oxygen saturation.  Offshore of this region is water that is very warm and salty in comparison ($\gamma > 0.15 \mathrm{kg\,m^{-3}}$), and relatively high in oxygen (saturations of approximately 60\%).  The region between these two water masses is very abrupt, and  stretches from La Perouse Bank to a shallow bank just above the Juan de Fuca canyon (48.3\textdegree N and 125.4\textdegree W).

@@

\subsubsection{Spur Canyon}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The Spur Canyon leading from the Strait of Juan de Fuca has been implicated in allowing dense water to be upwelled into the \Eddy.  The sea surface is low in the middle of the \Eddy, so it has been hypothesized that water moves up the Spur Canyon due to ageostrophic motion \cite{weaverhsieh87,freelanddenman82}.  This is difficult to infer from the observations collected here.  Three transects up the canyon indicate that deep isopycnals slope up into the canyon (\fref{fig:CanyonPropertiesSpice}, to approximately 30 km).  Continuing across the \Eddy, isopycnals are largely flat until they intersect the Vancouver Island Coastal Current ($80\ \mathrm{km}$).  The deeper isopycnals are not found in the deeper basin northeast of La Perouse bank ($70\ \mathrm{km}$),  thus the bank is a natural poleward boundary of the \Eddy.

@@

Much of the modified water in the Spur Canyon appears to come from the shelf to the west, but heavily modified by tidal mixing. A repeat tidal survey over the bank on the west side of the canyon shows a strong hydraulic response during onshore flow (17:58--21:00, \fref{fig:HydraulicsCanyon}).  Dense water passes from the offshore side into the canyon, and plunges down the side wall before rebounding downstream.  Note that the tide here is largely diurnal, so this onslope flow only occurs once a day.  Given the stratification of $N\approx 6\times10^{-3} \ \mathrm{rad\,s^{-1}}$ and an overturning scale of 50 m, we might expect dissipation rates reaching $\epsilon \sim L^2N^{3} \approx 5\times10^{-4}\ \mathrm{m^2\,s^{-3}}$, which is three orders of magnitude higher than dissipation observed on the shelf west of this location by \citeA{deweycrawford88}.  This estimate of turbulence dissipation rate implies a diapycnal diffusivity of $\kappa = \gamma \epsilon / N^2 \approx 1\ \mathrm{m^2\,s^{-1}}$, assuming a mixing efficiency of $\gamma=0.2$. Water spills over from the offshore front into the canyon during the onshore tide.  This water is rapidly mixed with surrounding water such that its strong offshore spice values are attenuated.

@@

The turbulence found on the canyon rim makes it ambiguous if there is water moving up the canyon or not.  There is a general tendency along the canyon for higher spice water to be found offshore (\fref{fig:CanyonPropertiesSpice}) but it seems likely that the source of the higher spice water is from over the bank rather than water being advected up the canyon.  This tidally driven flow over the bank is the most significant source of high-spice offshore water into the eddy region identified during our surveys.

\subsubsection{Frontal survey}
\label{sec:frontsurvey}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Separation of coastal water}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Upstream of the the sharp front between the low-spice \Eddy\ and the high-spice offshore water is a substantial mass of intermediate-spice water along the shelf (compare \fref{fig:CrossSectionsSpice}a and c).   This intermediate spice water is moving equatorward along the shelf upstream of the \Eddy, but is pushed offshore just upstream of Barkley Canyon (\fref{fig:IsopycnalsDepth}).  There is a tongue of intermediate-spice water ($\gamma$ is pink along $26.4\,\mathrm{kg\,m^{-3}}$) that separates from the shelf just west of 126\textdegree W. The surveys do not cross the full extent of the  tongue, but it is at least 30 km wide. It also appears to end at approximately 48.2\textdegree N.  This intermediate-spice water reaches from relatively shallow isopycnals to at least $26.6\,\mathrm{kg\,m^{-3}}$ (\fref{fig:IsopycnalsDepth}, left panel).  Unfortunately, we cannot track the fate of this water mass because isopycnals tilt down offshore, below the depth limit of the MVP.   The $25.5\,\mathrm{kg\,m^{-3}}$ isopycnal appears to have the tongue closer to the shelf than at $26.4\,\mathrm{kg\,m^{-3}}$, indicating strong three-dimensionality to this feature.
